\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid

### 2014-10-02 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Simple edits requested by Bill.

\begin{abstract}
%Should be about 200 words
The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
evolutionary optimization without requiring problem specific parameters.
This is achieved by combining local search with advanced crossover to iteratively construct
and exploit multiple populations simultaneously, with each population akin to a generation
in traditional search.

When compared with five advanced search algorithms across dozens of problem sizes for
seven different benchmarks P3 reliably required at least a constant factor less evaluations to find the global optimum
than all tested competitors. When compared using fixed numbers of evaluations P3
generally finds the highest quality solutions. Some evidence also suggests this performance
extends to considering total number seconds to find the global optimum even when using efficient fitness functions.
We investigate the causes of this efficiency and provide some evidence that P3's ability to
add diversity only as needed may be contributing to its success.

This quality is made even more striking in that unlike P3, the best comparison algorithms received extensive
problem specific tuning to achieve their performance. As a result we conclude that in contrast to
previous parameter-less techniques P3 is more efficient than existing parameterized methods for
black box optimization.

@@

In evolutionary optimization, and genetic algorithms~(GAs) in particular, search is often
biased through parameters. This can be beneficial as it allows practitioners to inject their
knowledge about the shape of the search landscape into the algorithm.
However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
correctly~\citep{goldberg:1991:gasize}. As such expert knowledge or exceedingly
expensive parameter tuning~\citep{grefenstette:1986:optimalga} may be required to leverage
this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
rate, tournament size, etc can have no clear relationship to the problem being solved, meaning even
domain experts may not understand how they will interact with the problem or with each other.

As such there have been periodic efforts to reduce or remove the need for parameter tuning.
\cite{Back:1992:selfadapt} introduced self-adaptive parameters, in which parameter values
were included in each solution's genome and underwent evolution. This allowed search
to optimize some of its own parameters resulting in a reduced the need for expert tuning.
\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.

One area that has been very effective at reducing the number of algorithm parameters is
model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
both only require a single parameter: population size. \cite{posik:2011:parameterless}
leveraged model building to create a fully parameter-less algorithm, but it is restricted to
only order-k, fully decomposable, noiseless problems.

@@

In order to fully understand the effectiveness of a black box search algorithm, it is useful
to compare it with other similar algorithms. Therefore here we preset five advanced algorithms with
related features to P3. The Random Restart Hill Climber was chosen as an efficient form of repeated
local search, and because it shows the advantages of performing the more complex portions of P3.
The $1+(\lambda, \lambda)$ algorithm is the current best theory supported simple genetic algorithm
and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
hBOA are advanced model building search techniques which are very effective at learning complex
problem structure, designed to achieve similar goals as P3's linkage learning but using very different
methods. Finally LTGA represents the current state-of-the-art in black box search and is the origin
of P3's linkage learning and crossover methods.

Finally, only hBOA and LTGA require any parameters, with each of these only requiring a population
size. This makes knowing the optimal behavior of these algorithms much more tractable. All of the
algorithms are also gene order independent, fitness scale invariant, and unbiased. This means
for any problem the order in which problem variables appear in the genome can be changed
without changing the behavior of the search. The fitness can also be manipulated in any fashion
as long as the rank ordering of solutions is unchanged. These algorithms are also unaffected by the meaning
assigned to each bit, such that inverting a predetermined random subset of genes before evaluation
will not impact search efficiency.

@@

\begin{figure}
  \begin{algorithmic}[1]
  \Procedure{Hill-Climber}{}
    \State $options \leftarrow [0 \dots N-1]$
    \State $tried \leftarrow \emptyset$
    \While{$|tried| < |options|$}
      \ForAll{$index \in shuffled(options)$}
        \If{$index \notin tried$}
          \State Flip bit $index$ in solution
          \If{solution's fitness increased}
            \State $tried \leftarrow \emptyset$
          \Else
            \State Revert change
          \EndIf
          \State $tried \leftarrow tried \cup \{index\}$
        \EndIf
      \EndFor
    \EndWhile
  \EndProcedure
\end{algorithmic}
  \caption{Hill climbing algorithm used to improve randomly generated solutions until no single
           bit change results in a fitness improvement.}
  \label{fig-hc}
\end{figure}

The hill climber requires an amortized cost of $O(1)$ operations per evaluation. In order to
terminate, at least one evaluation must be performed for each of the $N$ bits in the solution.
As such any operation that happens only once per search can be amortized over at least $N$
evaluations, covering the initialization of $options$.

With the use of proper hashing data structures, determining if $index \notin tried$ requires $O(1)$
time. While this operation may be called without a directly related evaluation, amortized cost
provides an upper bound. Consider that in order to add an $index$ into $tried$ an evaluation must
be performed. Lets say this evaluation pays for two checks of that $index$, one of which was used
during this loop. If no fitness improving move is found, the next time through the loop the second check
is used. A third check is only necessary if no fitness improving move was found for any index.
This is a contradiction as the loop terminates when no improving move was found. Therefore this check
can only be called twice per evaluation.

Due to its nature, this hill climber cannot escape basins of attraction. Once a solution is reached
such that none of the single bit neighbors are fitness improvements, search stops. As a result to
have any hope of solving multimodal problems this algorithm requires a restart mechanism. We have
chosen here to naively restart search from a random solution whenever a local optima is found. This
ensures that on all landscapes there is always a non-zero probability of search finding the global optimum.

@@

To perform replacement, hBOA uses restricted tournament replacement. After each new solution is generated
and evaluated, $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
Whichever of the $w$ solutions is the most genetically similar to the offspring is compared with the offspring.
If the fitness of the offspring is no worse, it replaces the genetically similar member of $w$, otherwise the
offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
solutions must compete on fitness.

@@

\subsection{Linkage Tree Genetic Algorithm}
~\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced
ability to preserve high fitness gene subsets, LTGA was able to outperform state of the art
GAs across many benchmarks. Since its first proposal, many variants of LTGA have been
proposed~\citep{goldman:2012:ltga} so for clarity we have chosen the version
presented by \cite{thierens:2013:ltgahiff} as our model. This variant was chosen as it is
modern and appears to approach the consensus in the literature.

@@

Figure~\ref{fig-cluster-creation} provides the agglomerative method LTGA uses to create gene clusters.
This algorithm will create a tree of sets, such that the leaves of the tree contain a single gene and
nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
relationships which should be preserved.
The process begins by creating the set of sets $unmerged$ which tracks all top level clusters. Initially
$unmerged$ contains single member sets for each gene. After each iteration the two sets with the minimum average pairwise
distance (given in Equation~\ref{eq-distance}) are merged to create a single cluster. This process is repeated
until only a single set remains in $unmerged$ which contains all of the genes in the genome.

@@

Throughout this process $useful$ tracks the set of all gene clusters which should be preserved for use by crossover.
This set begins with all genes in separate clusters, and each time a new cluster is created it is added to $useful$.
However, not all clusters are necessarily worth keeping. For instance, in all versions of LTGA the cluster
containing all genes is removed from $useful$ as preserving all genes during crossover can only create clones.
\cite{thierens:2013:ltgahiff} extended this removal to include any unsupported subsets. If the pairwise distance
between two clusters is 0, this means there are no individuals in the population which disrupt the relationships between the two
clusters. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
by breaking the stored pattern. As such a cluster is only kept if its direct superset has a non-zero distance.

@@

In order to efficiently perform clustering, a pairwise gene frequency table is constructed.
Equation~\ref{eq-entropy} uses how frequently each of the four possible string values for each
pair of genes occurs in the selected solutions. Extracting this information requires $O(\mu N^2)$
time, where $\mu$ is the population size and $N$ is the genome size. The process of converting
this pairwise frequency information into clusters can be achieved in $O(N^2)$ using the bookkeeping
methods presented by \cite{gronau:2007:upgma}. This cost is performed only once per generation,
and is then used to perform approximately $O(\mu N)$ crossover evaluations. As a result, the amortized cost of
LTGA's model building is $O(N)$.

@@

Each generation, each individual in the population undergoes crossover. In a single crossover event, each
cluster of genes $C_i$ in $useful$ is applied in last merged first order as a crossover mask. A random donor $d$
is chosen from the entire population (not just the model selected population), and $d$'s gene's for $C_i$ are copied
into the working solution. If a modification is made, an evaluation is then performed. If the crossover
resulted in no worse fitness then the changes are kept. This allows for neutral drift across plateaus.
The resulting solution, which must be at least as fit as its parent, is then
copied into the next generation.

@@

To accommodate P3's unique population structure, some of LTGA's clustering procedures were adapted. In LTGA,
clusters are identified at the start of each generation, with those clusters used
to create all solutions produced in that generation. As P3 does not perform serial
generations, this process had to be modified. Instead, P3 updates the model each
time a solution is added to a population. Furthermore, unlike
LTGA all solutions in the population are used to generate the model, not just the
winners of a binary tournament. This selection process is not necessary as the
worst solutions in the pyramid are already high quality due to local search.
Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
to provide no significant improvement. A likely cause was that that study applied
local search to every solution, not just the initial population, resulting in significant overhead.

Beyond the changes in population structuring, P3 also has slight modifications to
LTGA's version of \Call{Cluster-Creation}{} and \Call{Cluster-Usage}{}.
P3 changes Line~\ref{fig-cluster-creation-ordering} in Figure~\ref{fig-cluster-creation}
from \emph{last merged first} ordering to \emph{smallest first} ordering.
This method applies gene clusters during crossover based on how many genes
are in each cluster\footnote{Ties are broken randomly.}, and not on how tightly linked those genes are.
\cite{goldman:2012:ltga} found that this alternative was better at preserving
diversity, and therefore required smaller populations, but has  not been integrated into
our chosen canonical LTGA variant.

@@

Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
and only If (HIFF) problem is designed to capture the difficulties of this class of problem.
In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
leaf and each internal node is the subset of genes contained in its children. If all genes represented
in a node of the tree are set to the same value, they score equal to the size of the set. In this way
small subsets lead toward solutions to larger subsets. However a node can score if all genes are either
all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
perfectly duplicate the problem's true relationships \citep{thierens:2013:ltgahiff}.

@@

\subsection{Randomly Generated Problem Classes}
While well defined landscapes can provide specific insights into how an algorithm works,
their static nature can create misleading results. Specifically algorithm quality might
be very fragile such that it is only effective at searching well behaved landscapes. A more realistic
test of an algorithm's black box effectiveness is to work with randomly generated instances
drawn from a problem class. When tested over a sufficiently large sample it is then possible
to draw more general conclusions about the algorithms effectiveness. The challenge with these
landscapes is determining the global optimum to know if an algorithm was successful.

Perhaps the most common model for generating random rugged landscapes is the NK model. An
NK Landscape determines the fitness of each gene based on epistatic relationships with $K$
other genes in the genome. This fitness is specified using a randomly generated table of
fitness values, were each possible combination of the $K+1$ genes is mapped to some floating
point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
if epistasis is instead set such that each gene depends on the $K$ directly following it
in the genome the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.

\cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
Ising Spin Glasses. A spin glass is defined by interaction terms between variables with
the genome encoding the sign for each variable and a goal to minimize the total interaction value.
Similar to NK Landscapes the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}

As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
Related to the more common 3-SAT problem, a MAX-SAT instance is defined by a set of three term clauses.
Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
optimum \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
problem is challenging chose a clause to variable ratio of $4.27$.

\section{Comparison Algorithm Parameter Tuning}
\label{sec-tuning}
While four of the six algorithms in our experiments do not require any user specified parameters,
hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
handicapped, we have extensively tuned each using the bisection method \citep{sastry:2001:bisection}
to determine the optimal population size for each size of each problem.
Extended by \cite{goldman:2012:ltga} this method iteratively
doubles the population size until some success criteria is met and then performs bisection
between the lowest successful and highest unsuccessful sizes. Thus the minimum population size
which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
As bisection can make infinitely large population sizes, any run which had not found the global
optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.

@@

As expected, Parameter-less hBOA was generally less efficient than hBOA using an optimal population size. The slopes
of the two lines are relatively similar across all problems, meeting with the theoretical expectation that
Parameter-less hBOA is no more than a logarithmic factor worse than hBOA.  The only problem which reversed
this trend was MAX-SAT, where Parameter-less hBOA actually outperformed hBOA. A potential explanation for
this upset is that hBOA is using the optimal population size for the problem class, not for each instance.
Therefore hBOA must use a population size sufficiently large to solve the hardest instances in the class.
Parameter-less hBOA on the other hand is able to solve easy instances using much smaller population sizes.
The net effect is therefore that by being more adaptable Parameter-less hBOA is able to achieve better median
results.

@@

Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
The largest of these used 49 bits, with P3 achieving the best results on all 92 larger
configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the 5, 4, and 1 smallest Deceptive
Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the 3 smallest Nearest Neighbor NK instances
and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
on 5 smallest Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.

@@

For some applications, finding the global optimum is less important than finding good solutions quickly. Therefore
we examine this behavior in Figure~\ref{fig-fitness-over-time}. At regular intervals during optimization
Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found by that point of search across
100 runs. For each problem we show the largest problem size for which we were able to successfully set a population
size for hBOA, but the trends shown are representative of all larger problem sizes. The maximum reporting interval
is set to include the slowest P3 run to reach the global optimum.

@@

At 63\% of recording intervals hBOA has the worst fitness of any solver. Many of the exception occur when
the total number of evaluations performed is significantly less than the genome size and hBOA reaches better fitness
than the Hill Climber, $1+(\lambda, \lambda)$, and P3. This early in search hBOA and LTGA are both still evaluating
individuals in their randomly generated initial populations, while the three techniques worse than hBOA have started
from a single solution and are attempting to optimize. Of the methods, hBOA uses by far the largest population sizes,
about 55 times larger than LTGA on average for problems shown. As such it makes sense that hBOA would take a long time to reach high
quality solutions.

>>>>

\begin{abstract}
%Should be about 200 words
The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
evolutionary optimization without requiring any user specified parameters.
This is achieved by combining local search with advanced crossover to iteratively construct
and exploit multiple populations simultaneously, with each population akin to a generation
in traditional evolutionary computation.

When compared with five advanced search algorithms across dozens of problem sizes for
seven different benchmarks, P3 reliably required at least a constant factor less evaluations to find the global optimum
than all tested competitors. When compared using fixed numbers of evaluations, P3
generally finds the highest quality solutions. Some evidence also suggests that P3
is better when compared using the number seconds to find the global optimum even when using efficient fitness functions.
We investigate the causes of this efficiency and provide some evidence that P3's ability to
add diversity only as needed may be contributing to its success.

This quality is made even more striking in that unlike P3, the best comparison algorithms received extensive
problem specific tuning to achieve their performance. As a result we conclude that, in contrast to
previous parameter-less techniques, P3 is an efficient, general, parameter-less approach to black box
optimization which is more effective than existing state-of-the-art techniques.

@@

In evolutionary optimization, and genetic algorithms~(GAs) in particular, search is often
biased through parameters. This can be beneficial as it allows practitioners to inject their
knowledge about the shape of the search landscape into the algorithm.
However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
correctly~\citep{goldberg:1991:gasize}. As such, expert knowledge or exceedingly
expensive parameter tuning~\citep{grefenstette:1986:optimalga} may be required to leverage
this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
rate, tournament size, etc. may have no clear relationship to the problem being solved, meaning even
domain experts may not understand how they will interact with the problem or with each other.

There have been periodic efforts to reduce or remove the need for parameter tuning.
\cite{Back:1992:selfadapt} introduced self-adaptive parameters, in which parameter values
were included in each solution's genome and themselves underwent evolution. This allowed search
to optimize some of its own parameters resulting in a reduced the need for expert tuning.
\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.

One area that has been very effective at reducing the number of algorithm parameters is
model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
both require only a single parameter: population size. \cite{posik:2011:parameterless}
leveraged model building to create a fully parameter-less algorithm, but it is restricted to
only order-k, fully decomposable, noiseless problems.

@@

In order to fully understand the effectiveness of P3, we compare it with five advanced algorithms which
have related features to P3. The Random Restart Hill Climber~\citep{goldman:2014:p3} was chosen as an efficient form of repeated
local search. As P3 combines this hill climber with crossover,
comparing with local search alone shows the advantages of performing the other portions of P3.
The $1+(\lambda, \lambda)$ algorithm is the current best theory supported simple genetic algorithm
and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
hBOA are advanced model building search techniques which are very effective at learning complex
problem structure, designed to achieve similar goals as P3's linkage learning but using very different
methods. Finally LTGA represents the current state-of-the-art in black box search and is the origin
of P3's linkage learning and crossover methods.

Finally, only hBOA and LTGA require any parameters, with each of these only requiring a population
size. This makes knowing the optimal behavior of these algorithms much more tractable. All of the
algorithms are also gene order independent, fitness scale invariant, and unbiased. This means,
for any problem, the order in which problem variables appear in the genome can be changed
without changing the behavior of the search. The fitness can also be manipulated in any fashion
as long as the rank ordering of solutions is unchanged. These algorithms are also unaffected by the meaning
assigned to each bit, such that inverting a predetermined random subset of genes before evaluation
will not impact search efficiency.

@@

\begin{figure}
  \begin{algorithmic}[1]
  \Procedure{Hill-Climber}{}
    \State $options \leftarrow [0 \dots N-1]$\label{fig-hc-options}
    \State $tried \leftarrow \emptyset$
    \While{$|tried| < |options|$}
      \ForAll{$index \in shuffled(options)$}
        \If{$index \notin tried$}
          \State Flip bit $index$ in solution
          \If{solution's fitness increased}
            \State $tried \leftarrow \emptyset$
          \Else
            \State Revert change
          \EndIf
          \State $tried \leftarrow tried \cup \{index\}$
        \EndIf
      \EndFor
    \EndWhile
  \EndProcedure
\end{algorithmic}
  \caption{Hill climbing algorithm used to improve randomly generated solutions until no single
           bit change results in a fitness improvement.}
  \label{fig-hc}
\end{figure}

The hill climber requires an amortized cost of $O(1)$ operations per evaluation. In order to
terminate, at least one evaluation must be performed for each of the $N$ bits in the solution.
As such any operation that happens only once per search can be amortized over at least $N$
evaluations, covering the initialization of $options$ on Line~\ref{fig-hc-options}.

With the use of proper hashing data structures, determining if $index \notin tried$ requires $O(1)$
time. While this operation may be called without a directly related evaluation, amortized cost
provides an upper bound. Consider that in order to add an $index$ into $tried$ an evaluation must
be performed. Lets say this evaluation pays for two checks of that $index$, one of which was used
during this loop. If no fitness improving move is found, the next time through the loop the second check
is used. A third check is only necessary if no fitness improving move was found for any index.
This is a contradiction as the loop terminates when no improving move was found. Therefore this check
is called at most twice per evaluation.

Due to its nature, this hill climber cannot escape basins of attraction. Once a solution is reached
such that none of the single bit neighbors are fitness improvements, search stops. Thus
this algorithm requires a restart mechanism to solve multimodal problems. We have
chosen here to naively restart search from a random solution whenever a local optima is found. This
ensures that on all landscapes there is always a non-zero probability of search finding the global optimum.

@@

To perform replacement, hBOA uses restricted tournament replacement. After each new solution is generated
and evaluated, $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
Whichever of the $w$ solutions is the most genetically similar to the offspring compares its fitness with the offspring.
If the fitness of the offspring is no worse, it replaces the genetically similar member of $w$, otherwise the
offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
solutions must compete on fitness.

@@

\subsection{Linkage Tree Genetic Algorithm}
~\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced
ability to preserve high fitness gene subsets, LTGA was able to outperform state of the art
GAs across many benchmarks. Since its introduction, many variants of LTGA have been
proposed~\citep{goldman:2012:ltga} so for clarity we have chosen the version
presented by \cite{thierens:2013:ltgahiff} as our model. This variant was chosen as it is
modern and appears to approach the consensus in the literature.

@@

Figure~\ref{fig-cluster-creation} provides the agglomerative method LTGA uses to create gene clusters.
This algorithm creates a tree of sets using pairwise gene entropy, such that the leaves of the tree contain a single gene and
internal nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
relationships which should be preserved.
The process begins by creating the set of sets $unmerged$ which tracks all top level clusters. Initially
$unmerged$ contains single member sets for each gene. After each iteration the two sets with the minimum average pairwise
distance (given in Equation~\ref{eq-distance}) are merged to create a single cluster. This process is repeated
until only a single set remains in $unmerged$ which contains all of the genes in the genome.

@@

Throughout this process $useful$ tracks the set of all gene clusters which should be preserved for use by crossover.
This set begins with all genes in separate clusters, and each time a new cluster is created it is added to $useful$.
However, not all clusters are necessarily worth keeping. For instance, in all versions of LTGA the cluster
containing all genes is removed from $useful$ as preserving all genes during crossover can only create clones.
\cite{thierens:2013:ltgahiff} extended this removal to include any unsupported subsets. If the pairwise distance
between two clusters is 0, this means there are no individuals in the population which disrupt the relationships between the two
clusters. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
by breaking the stored pattern. As such a cluster is only kept if its direct superset has a non-zero distance.
As a final step Line~\ref{fig-cluster-creation-ordering} reorders $useful$ such that clusters appear in reversed
order from which they were added to $useful$. Thus the most linked clusters and those containing single genes
appear at the end of the returned list.

@@

In order to efficiently perform clustering, a pairwise gene frequency table is constructed from the selected solutions.
To calculate Equation~\ref{eq-distance}, Equation~\ref{eq-entropy} is called for each
gene ($H(c_i)$) and pair of genes ($H(c_i \cup c_j)$). Extracting this information requires $O(\mu N^2)$
time, where $\mu$ is the population size and $N$ is the genome size. The process of converting
this pairwise frequency information into clusters can be achieved in $O(N^2)$ using the bookkeeping
methods presented by \cite{gronau:2007:upgma}. This cost is performed only once per generation,
and is then used to perform approximately $O(\mu N)$ crossover evaluations. As a result, the amortized cost of
LTGA's model building is $O(N)$.

@@

Each generation, each individual in the population undergoes crossover. In a single crossover event, each
cluster of genes $C_i$ in $useful$ is applied as a crossover mask. A random donor $d$
is chosen from the entire population (not just the model selected population), and $d$'s gene's for $C_i$ are copied
into the working solution. If a modification is made, an evaluation is then performed. If the crossover
resulted in no worse fitness then the changes are kept. This allows for neutral drift across plateaus.
The resulting solution, which must be at least as fit as its parent, is then
copied into the next generation.

@@

To accommodate P3's unique population structure, some of LTGA's clustering procedures were adapted. In LTGA,
clusters are identified at the start of each generation, with those clusters used
to create all solutions produced in that generation. As P3 does not perform serial
generations, this process had to be modified. Instead, P3 updates the model each
time a solution is added to a population. Furthermore, unlike
LTGA all solutions in the population are used to generate the model, not just the
winners of a binary tournament. We can do this because even the
worst solutions in the pyramid are already high quality due to local search.
Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
to provide no significant improvement. A likely cause was that that study applied
local search to every solution, not just the initial population, resulting in significant overhead.

Beyond the changes in population structuring, P3 also has slight modifications to
LTGA's version of \Call{Cluster-Creation}{} and \Call{Cluster-Usage}{}.
P3 changes Line~\ref{fig-cluster-creation-ordering} in Figure~\ref{fig-cluster-creation}
from \emph{last merged first} ordering to \emph{smallest first} ordering.
This method applies gene clusters during crossover based on how many genes
are in each cluster\footnote{Ties are broken randomly.}, and not on how tightly linked those genes are.
\cite{goldman:2012:ltga} found that this alternative was better at preserving
diversity, and therefore required smaller populations, but has  not been integrated into
the canonical LTGA variant we used.

@@

Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
and only If (HIFF) problem is designed to capture the difficulties of this class of problem.
In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
leaf and each internal node is the subset of genes contained in its children. If all genes represented
in a node of the tree are set to the same value, they score equal to the size of the set. In this way
small subsets lead toward solutions to larger subsets. However, a node can score if all genes are either
all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
perfectly duplicate the problem's true relationships \citep{thierens:2013:ltgahiff}.

@@

\subsection{Randomly Generated Problem Classes}
While well defined landscapes can provide specific insights into how an algorithm works,
their static nature can be misleading. Specifically, algorithm quality might
be very fragile such that it is only effective at searching well behaved landscapes. A more realistic
test of an algorithm's black box effectiveness is to work with randomly generated instances
drawn from a problem class. When tested over a sufficiently large sample it is then possible
to draw more general conclusions about the algorithms effectiveness. The challenge with these
landscapes is determining the global optimum to gauge if an algorithm was successful.

Perhaps the most common model for generating random rugged landscapes is the NK model. An
NK Landscape determines the fitness of each gene based on epistatic relationships with $K$
other genes in the genome. This fitness is specified using a randomly generated table of
fitness values, were each possible combination of the $K+1$ genes is mapped to some floating
point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
if epistasis is set such that each gene depends on the $K$ directly following it
in the genome the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.

\cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
Ising Spin Glasses. A spin glass is defined by a weighted graph of interaction terms between vertices.
Each gene assigns a value to each vertex, with the fitness calculated by Equation~\ref{eq-ising}.
In this equation, $E$ is the set of all edges, $e_{ij}$ is the edge weight connecting vertex $i$ to vertex $j$, and
$x_i$ and $x_j$ are the gene values for vertex $i$ and $j$. Optimal fitness is when this sum is minimized.
\begin{equation}
\sum_{e_{ij} \in E} x_ie_{ij}x_j
  \label{eq-ising}
\end{equation}
Similar to NK Landscapes, the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
In this subset the graph is restricted to be a two dimensional torus, edge weights are randomly set
to either -1 or 1, and vertex values must be -1 or 1.

As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
Related to the more common 3-SAT problem, a MAX-SAT instance is defined by a set of three term clauses.
Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
optimum, \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
problem is challenging chose a clause to variable ratio of $4.27$.

\section{Comparison Algorithm Parameter Tuning}
\label{sec-tuning}
While four of the six algorithms in our experiments do not require any user specified parameters,
hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
handicapped, we have extensively tuned each using the bisection method \citep{sastry:2001:bisection}
to determine the optimal population size for each size of each problem.
Extended by \cite{goldman:2012:ltga}, this method iteratively
doubles the population size until some success criteria is met and then performs bisection
between the lowest successful and highest unsuccessful sizes. Thus the minimum population size
which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
As bisection can make infinitely large population sizes, any run which had not found the global
optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.

@@

As expected, Parameter-less hBOA was generally less efficient than hBOA using an optimal population size. The slopes
of the two lines are relatively similar across all problems, meeting with the theoretical expectation that
Parameter-less hBOA is no more than a logarithmic factor worse than hBOA.  The only problem which reversed
this trend was MAX-SAT, where Parameter-less hBOA actually outperformed hBOA. A potential explanation for
this is that hBOA is using the optimal population size for the problem class, not for each instance.
Therefore hBOA must use a population size sufficiently large to solve the hardest instances in the class.
Parameter-less hBOA on the other hand is able to solve easy instances using much smaller population sizes.
The net effect is therefore that by being more adaptable Parameter-less hBOA is able to achieve better median
results.

@@

Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
The largest of these used 49 bits, with P3 achieving the best results on all 92 larger
configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.

@@

For some applications, finding the global optimum is less important than finding good solutions quickly. Therefore
we examine this behavior in Figure~\ref{fig-fitness-over-time}. At regular intervals during optimization
Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found at that point of search across
100 runs. For each problem we show the largest problem size for which we were able to successfully set a population
size for hBOA, but the trends shown are representative of all larger problem sizes. The maximum reporting interval
is set to include the slowest P3 run to reach the global optimum.

@@

At 63\% of the recording intervals hBOA has the worst fitness of any solver. Many of the exceptions occur when
the total number of evaluations performed is significantly less than the genome size. Under these circumstances hBOA reaches better fitness
than the Hill Climber, $1+(\lambda, \lambda)$, and P3. Early in search hBOA and LTGA are both still evaluating
individuals in their randomly generated initial populations, while the three techniques worse than hBOA have started
from a single solution and are attempting to optimize. Of the methods, hBOA uses by far the largest population sizes,
about 55 times larger than LTGA on average for problems shown. As such it makes sense that hBOA would take a long time to reach high
quality solutions.
